### 2015-02-16 paper/stmts.tex
statement preservation and progress

\subsection*{Statement Preservation}
\subsection*{Statement Progress}
\section*{Patina Soundness}

>>>>

\subsection*{Statement Preservation}
Statement preservation is the assertion that given all our well-formed contexts,
a well-typed statement, and the fact that that statement evaluates to another statement,
then we can also type the result statement and both statements produce the same output
shadow heap, i.e.\ they effectively end in the same way.
Put another way, this lemma asserts that memory safe programs step to memory safe programs.

It is difficult to specify how the new versions of contexts relate to the old versions.
They may be larger or smaller depending upon the statement in question.
We do not specify this relationship currently.
Instead we just rely on the new variable map, heap and statement to enforce
the relationship.

\begin{conj}[Statement Preservation]
  If $\tc{}{H}{\gS}$, $\gG;V\vdash\gS$, $\vdash\ltr$, $\tb{\ltr}{\gU}{\gG}{L}$,
  $\tc{V}{H}{\gU}$, $\tc{\gG;\ltr;L;\gU}{s}{\gU_0}$, and $\sev{V}{H}{s}{V'}{H'}{s'}$
  then there exist $\gS'$, $\gG'$, $\ltr'$, $L'$, and $\gU'$ such that
  $\tc{}{H'}{\gS'}$, $\gG';V'\vdash\gS'$, $\vdash\ltr'$, $\tb{\ltr'}{\gU'}{\gG'}{L'}$,
  $\tc{V'}{H'}{\gU'}$, and $\tc{\gG';\ltr';L';\gU'}{s'}{\gU_0}$.

\end{conj}

\begin{proof}[Justification]
  The skip-seq case is trivial.
  The step-seq case follows from induction.
  In the value initialization (set) case, only the shadow heap and the heap are changing.
  From Expression Preservation, we know that the heap and shadow heap are still aligned
  after evaluating the expression. The \textsc{write} will initialize the path, which
  is reflected in the shadow heap.
  The pointer initialization (new) case is similar.
  Since the components of a swap must have the same shadow, the modified heap will
  still be well-formed.
  We required that the memory deallocated by \textsf{free} was droppable, which
  should mean that the heap is still well-formed once we $\void$ the pointer.
  Push steps immediately to pop and only extends contexts, which should still be well-formed.
  The pop-step case follows from induction.
  In the pop-skip case, similar to the \textsf{free} case, we know that the variable
  is droppable, which should mean the resulting contexts are well-formed.
\end{proof}

\subsection*{Statement Progress}
Statement progress is essentially the assertion that a well-typed statement will
never attempt to read uninitialized memory, free borrowed memory, and so on.
Essentially, that well-typed programs have sufficient memory safety to take a step.
The restriction to statements that end with an empty shadow heap allows us
to avoid the cases with orphan memory.

\begin{conj}[Statement Progress]
  If $\tc{}{H}{\gS}$, $\gG;V\vdash\gS$, $\vdash\ltr$, $\tb{\ltr}{\gU}{\gG}{L}$,
  $\tc{V}{H}{\gU}$, $\tc{\gG;\ltr;L;\gU}{s}{\emptyset}$
  then there exists $V'$, $H'$, and $s'$ such that $\sev{V}{H}{s}{V'}{H'}{s'}$
  or $s = \nop$.
\end{conj}

\begin{proof}[Justification]
  The skip case is immediate.
  The sequence case is either immediate or follows from induction.
  The swap case and the new case follow from path progress and read safety.
  The set case is similar, but also requires expression progress.
  The push case is immediate.
  The pop case is either immediate or follows from induction.
\end{proof}

\section*{Patina Soundness}
If our conjectures hold, then we have established soundness for Patina.
This means that well-typed statement that ends with an empty shadow heap
will not leak memory, not access uninitialized memory, not create aliased, mutable memory,
and so on. In effect, well-typed programs are statically guaranteed to be memory safe.
